Exact Phylodynamics via Structured Markov Genealogy Processes

### 2023-07-11 ms.Rnw
thms 1 & 2

Denote the underlying probability space by $(\Omega,\Borel,\mathbb{P})$.
We will assume that our population process is a time-inhomogeneous Markov jump process, $\X_t$, parameterized by time $t\in\Rp\coloneq\{t\in\mathbb{R}\;\vert\;t\ge{0}\}$ and taking values in some space $\Xspace$.
In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
The population process is completely specified by its initial-state distribution, $p_0$, and its transition rates $\alpha$.
In particular, we suppose that
\begin{equation}\label{eq:ic}
  \Prob{\X_0=x}=p_0(x),
\end{equation}
for some choice of initial distribution, $p_0$.
For any $t\in\Rp$, $x,x'\in\Xspace$, we think of the quantity $\alpha(t,x,x')$ as the instantaneous hazard of a jump from $x$ to $x'$.
More precisely, the transition rates have the following properties:
\begin{equation*}
  \begin{gathered}
    \alpha(t,x,x')\ge{0}, \qquad \int_{\Xspace}{\alpha(t,x,x')\,\dd{x'}}<\infty,\\
  \end{gathered}
\end{equation*}
for all $t\in\Rp$ and $x,x'\in\Xspace$.
Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
Let $\N_t$ be the number of jumps that $\X$ has taken by time $t$.
We assume that $\N_t$ is a simple counting process so that
\begin{equation*}
  \begin{gathered}
    \Prob{\N_{t+\Delta}=n+1\;\vert\;\N_{t}=n}=\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),
    \qquad
    \Prob{\N_{t+\Delta}>n+1\;\vert\;\N_{t}=n}=o(\Delta),\\
    \Prob{\X_{t+\Delta}\in\mathcal{E}\;\vert\;\X_{t}=x, \N_{t+\Delta}-\N_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}+\frac{o(\Delta)}{\Delta}.
  \end{gathered}
\end{equation*}
\AAK{[Do we need the last term?]}
We will further assume that $\X_t$ is non-explosive, \ie, that $\Prob{\N_t<\infty}=1$ for all $t$.
\AAK{[Is this equivalent to non-explosivity? Or merely an implication?]}

@@

Without loss of generality, one can assume, as we do here, that the sample paths $t\mapsto\X_t(\omega)$ for $\omega\in\Omega$ are right-continuous with left limits (\cadlag).
In fact, all of the processes we will describe in this paper will be taken to be \cadlag, and we will frequently need to refer to the left-limit.
Accordingly, if $z=z(\omega,t)$ is any \cadlag\ random function, we define
\begin{equation*}
  \zt(\omega,t)\coloneq
  \begin{cases}
    \lim_{t'\uparrow{t}}z(\omega,t'), & t>0,\\
    z(\omega,0), & t=0.\\
  \end{cases}
\end{equation*}
Note that $\zt$ is thus left-continuous with right limits.

@@

The non-explosiveness assumption implies that a.s., for every $t$, there is a finite, increasing sequence of random jump times $T_t\coloneq\left(t_k\right)_{k=1}^{K_t}$;
the length $K_t$ of this sequence is itself random.
However, conditional on $\Hist_t$, $T_t$ and $K_t$, together with the mark process $\U_t$, and the population process $\X_t$ are deterministic.
Using these, one can write down an explicit probability measure for $\Hist_t$:
\begin{equation}\label{eq:Hdens}
  \begin{aligned}
    \pi^{\Hist}(\dd{\Hist_t})
    =&\pi^{\Hist}(\dd{T_t},\dd{\X_t},K_t,\U_t)\\
    =&p_{0}(\X_{0})\,\dd{x_0}\,\prod_{k=1}^{K_{t}}{\left(\alpha_{\U_{t_k}}\left(t_k,\Xt_{t_k},\X_{t_k}\right)\,\dd{x_k}\,\dd{t_k}\right)}\\
    &\qquad\times\exp{\left(-\sum_{u}{\int_{0}^{t}{\int{\alpha_u(t',\X_{t'},x')\dd{x'}}\dd{t'}}}\right)}.
  \end{aligned}
\end{equation}

@@

Our goal in this paper is to probabilistically characterize how the genealogical relationships among lineages evolve through time.
Accordingly, we develop some notation for this purpose.
To begin with, we assign to each lineage a unique number $j\in\Zp$.
This can be done in any fashion, so long as no two lineages ever receive the same number.
For example, when a new lineage arises, we can assign it the smallest integer that has not yet been assigned.
We will define the \emph{inventory process}, $\Inv_t$, so that, for every lineage $j\in\Zp$, $\Inv_t(j)$ is the deme in which $j$ is found.
However, when $t$ is before the birth or after the death of lineage $j$, then clearly $\Inv_t(j)\notin\Demes$.
We say in this case that lineage $j$ is in the \emph{underdeme}, which we denote using the symbol $\eth$, so that we can write $\Inv_t(j)=\eth$ and define $\Demesplus\coloneq\Demes\cup\{\eth\}$.
With this definition, $\Inv:\Rp\times\Zp\to\Demesplus$.

@@

Different kinds of events that occur for the population process can have different kinds of effects on the inventory process, and indeed not every jump affects $\Inv_t$ at all.
From the point of view of the inventory process, there are five distinct \emph{types} of jumps, which we enumerate here.
\begin{enumerate}[1.]
\item Birth-type events result in the branching of one or more new lineages, each from some existing lineage.
  If $j$ is one of the new lineages, we use the expression $\Anc(j)$ to refer to its ancestor.
  Examples of birth-type events include transmission events, speciations, and actual births.
  It is not assumed that all new lineages arising from a birth event share the same ancestor.
\item Death-type events result in the extinction of one or more lineages.
  Examples include recovery, death of a host, and species extinctions.
\item Migration-type events result in the movement of a lineage from one deme to another.
  Spatial movements, changes in behavior, and progression of an infection can all be represented as migration-type events.
\item Sample-type events result in the collection of a sample from a lineage but do not in themselves affect the inventory process.
\item Neutral-type events result in no change to any of the lineages.
\end{enumerate}
\Cref{fig:markov_state} depicts an example with all five of these types.
It is not necessary that a jump fall into just one of these types.
It is allowable, for instance to have compound jumps that fall into more than one category.
For example, Sample-death-type events, in which a lineage is simultaneously sampled and removed, have been used, as have birth-death events in which one lineage reproduces at the same moment that another dies.
The theory presented here places no restrictions on these events.

However, we do impose the restriction that the \emph{production}, \ie the deme-specific number of lineages emerging from the event, be constant for all jumps of a given mark.
To be precise, the \emph{production} is defined to be a function $r:\Jumps\times\Demes\to\Zp$, such that $r_u^i$ lineages of deme $i$ emerge from each event of mark $u$.
We write $r_u=(r_u^i)_{i\in\Demes}$.
Note that we do not count lineages that die as a result of the event.
Also, it is important to note that the parent lineage or lineages, if they survive the event, are always counted among the emergent lineages.

Because different kinds of events may differ not only in the number of offspring they engender, but also in the number of parent lineages, and the distribution of offspring among parents, there is also assumed to be a deterministic function $Q_u$, for $u\in\Jumps$, (described below) that expresses these properties.

@@

The structure of the state space for the inventory process, $\Inv_t$, has already been described.
It remains to define its stochastic dynamics.
The $\Inv_t$ process follows the population process $\X_t$ in that jumps in $\Inv_t$ do not occur except when jumps in $\X_t$ occur: $\predec{\Inv}_t\ne\Inv_t$ implies $\predec{\N}_t\ne\N_t$.

At jumps of birth type and mark $u$, one or more random parents are selected from the appropriate deme(s).
The appropriate number of offspring are created in each deme.

@@

\citet{King2022}, showed how an unstructured population process induces a process on the space of genealogies.
Although we now treat a more general case, the construction is much the same, so we abbreviate the presentation.
Readers wishing for more detail should consult the earlier paper \citep{King2022}.

Formally, we define a \emph{genealogy}, $\Gen$, to be a finite sequence of \emph{internal nodes}, together with a time.
The time is denoted $T(\Gen)$ and the number of nodes is $K(\Gen)$.
The $k$-th node is $\Node{p}_k(\Gen)$ and we write $\Node{p}\in\Gen$ if $\Node{p}$ is one of the nodes of $\Gen$.
Each $\Node{p}\in\Gen$ has a creation-time, $T(\Node{p})$ and a parent, $\Anc(\Node{p})$.
%% The nodes are ordered so that $k<k'$ implies that $T(\Node{p}_k(\Gen))\le{T(\Node{p}_{k'}(\Gen))}$.
Root nodes are distinguished by being their own parents:
$\Node{p}$ is a root if and only if $\Anc(\Node{p})=\Node{p}$.
Every node also has one or more descendants called \emph{children}.
There are three kinds of children:
\begin{inparaenum}[(i)]
\item \emph{Internal nodes}.
  Children nodes must always be later than their parents:
  $T(\Anc(\Node{p}))<T(\Node{p})$.
\item \emph{Tip nodes}, which represent extant lineages.
  In particular, every lineage alive at time $t$ is the child of some node.
\item \emph{Sample nodes}.
  When a sample is collected at time $t$, a new node, $\Node{p}$ is added with $T(\Node{p})=t$ and the sample as child.
\end{inparaenum}

@@

The population process induces a stochastic process on the space of genealogies.
In particular, at each event in the population process, one or more of the following changes happen to the genealogy, according to the type of the event:
\begin{enumerate}[(a)]
\item A birth-type event at time $t$ results in the creation of one new internal node for each parent lineage.
  In particular, if $j$ is one of the parent lineages, and $\Node{p}$ is the new node, then $T(\Node{p})=t$, $\Anc(\Node{p})$ is the node in which $j$ lay prior to the event.
  The children of $\Node{p}$ include all the new lineages that branched from $j$, as well as $j$ itself.
\item In a death-type event, all the lineages $j$ that die are removed.
  Nodes without children are then recursively removed.
\item In a migration-type event, one node is added for each migrating lineage;
  each one takes the migrating lineage as child.
  The ancestor of the new node is the node in which the migrating lineage lay before the event.
\item At a sample-type event, one new node is introduced for each sampled lineage.
  Each one takes the sampled lineage as child, along with the sample node.
  The ancestor of the node is that in which the sampled lineage lay before the vent.
\item At a neutral-type event, no change is made to the genealogy.
\item Finally, events of compound type are readily accommodated by combining the rules just stated.
\end{enumerate}

@@

\paragraph{Lineages, event-counter, deme-residence}

Let $\Lin=\Lin(\Gen)$ denote the finite set of all samples represented in genealogy $\Gen$.
Let $\prec2$ be any ordering of $\Lin$ that is compatible with ancestry.
That is, if $j,j'\in\Lin$ are such that $j$ is ancestral to $j'$, then we must have $j\prec2{j'}$.
Using this ordering, we can uniquely associate each point on a genealogical tree with the least of those lineages that descend from that point.
In particular, any lineage $j\in\Lin$, corresponding to a sample taken at time $t^s_j$, can be traced backward from node to node until either it coalesces with some lesser lineage at some time $t^o_j>0$ or a root is reached (in which case, we define $t^o_j=0$).
Each node encountered along the way represents a genealogical event from which $j$ emerges.
Moreover, at each time $t\in\halfopen{t^o_j,t^s_j}$, lineage $j$ is in precisely one of the demes $\Demes$.
However, for $t\notin\halfopen{t^o_j,t^s_j}$, lineage $j$ does not exist.
To express this, we again say that lineage $j$ is in the \emph{underdeme}, which we denote using the symbol $\eth$.
We define $\Demesplus\coloneq\Demes\cup\{\eth\}$.

It will be useful to define a function that captures all the relevant features of a pruned genealogy.
Accordingly, let $\Yspace=\Zp\times\Demesplus\times\Lin$ and define $y:\Rp\times\Lin\to\Yspace$ so that, for $j\in\Lin$ and $t\in\Rp$:
\begin{compactenum}[(a)]
\item $\ct(y_j(t))\in\Zp$ is a counting process which increases by 1 at each event along lineage $j$.
\item $\deme(y_j(t))\in\Demesplus$ indicates in which deme lineage $j$ lies at time $t$.
  In particular $\deme(y_j(t))=\eth$ if $t\notin\halfopen{t^o_j,t^s_j}$.
\item $\anc(y_j(t))\in\Lin$ indicates the lineage in which the ancestors of lineage $j$ are found.
  In particular, $\anc(y_j(t))=j$ for $t\ge{t^o_j}$, but $\anc(y_j(t))$ is well defined for all $j$ and $t$.
  \AAK{[Can define this in a way similar to that of the original Kingman's coalescent\dots.]}
\end{compactenum}
One can verify that $y$ is \cadlag.
Define $\yt(t)=\lim_{t'\uparrow{t}}y(t')$.

To visualize these functions, one can make a correspondence between demes and colors.
Then a pruned genealogy is visualized as a tree with colored branches.
Knowing the function $\deme(y)$ is equivalent to knowing the coloring, while $\ct(y)$ determines the locations of events in the genealogy and $\anc(y)$ determines the topology.
Note in particular that $y_j(t)\ne\yt_j(t)$ if and only if $t$ is the time of an event from which lineage $j$ emerges.

@@

In the following, we will find that we need to count the deme-specific numbers of lineages present at a given time.
Accordingly, for any $\Lin'\subseteq{\Lin}$, $\eta\in(\Zp\times\Demesplus\times\Lin')^{\Lin'}$, and $i\in\Demes$, let us define
\begin{equation*}
  \ell_i(\eta)\coloneq{\left\vert\left\{j\in\Lin'\;\vert\;\deme(\eta_j)=i\right\}\right\vert}\in\Zp, \qquad \ell(\eta)\coloneq(\ell_i(\eta))_{i\in\Demes}\in{\Zp^\Demes}.
\end{equation*}
Note that lineages $j$ for which $\eta_j=\eth$ are not counted.
With this definition, it follows that $\ell_i(y(t))$ is the number of lineages in deme $i$ at time $t$.

We will also have occasion to refer to the deme-specific number of lineages emerging from a given event.
Accordingly, for $\Lin'\subseteq{\Lin}$, $\eta,\eta'\in(\Zp\times\Demesplus\times\Lin')^{\Lin'}$, and $i\in\Demes$, let us define
\begin{equation*}
  s_i(\eta,\eta')\coloneq{\left\vert\left\{j\in{\Lin'}\;\vert\;\deme(\eta_j)=i\ \ \&\ \ \ct(\eta'_j)=\ct(\eta_j)+1\right\}\right\vert}, \qquad s(\eta,\eta')\coloneq(s_i(\eta,\eta'))_{i\in\Demes}\in{\Zp^\Demes}.
\end{equation*}
With this definition, $s_i(\yt(t),y(t))$ is the number of lineages in deme $i$ that emerge from an event at time $t$ and that, if $t$ is not an event time, then $s(\yt(t),y(t))=0$.

\subsection{Obscuration}

@@

\paragraph{Likelihood of an obscured genealogy}

>>>>

Denote the underlying probability space by $(\Omega,\Borel,\mathrm{Prob})$.
We will assume that our population process is a time-inhomogeneous Markov jump process, $\X_t$, parameterized by time $t\in\Rp\colonequals\{t\in\mathbb{R}\;\vert\;t\ge{0}\}$ and taking values in some space $\Xspace$.
In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
The population process is completely specified by its initial-state distribution, $p_0$, and its transition rates $\alpha$.
In particular, we suppose that
\begin{equation}\label{eq:ic}
  \Prob{\X_0\in\mathcal{E}}=\int_{\mathcal{E}}{p_0(x)\,\dd{x}}
\end{equation}
for all measurable sets $\mathcal{E}\subseteq\Xspace$.
For any $t\in\Rp$, $x,x'\in\Xspace$, we think of the quantity $\alpha(t,x,x')$ as the instantaneous hazard of a jump from $x$ to $x'$.
More precisely, the transition rates have the following properties:
\begin{equation*}
  \begin{gathered}
    \alpha(t,x,x')\ge{0}, \qquad \int_{\Xspace}{\alpha(t,x,x')\,\dd{x'}}<\infty,\\
  \end{gathered}
\end{equation*}
for all $t\in\Rp$ and $x,x'\in\Xspace$.
Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
Let $\N_t$ be the number of jumps that $\X$ has taken by time $t$.
We assume that $\N_t$ is a simple counting process so that
\begin{equation*}
  \begin{gathered}
    \Prob{\N_{t+\Delta}=n+1\;\vert\;\N_{t}=n}=\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),
    \qquad
    \Prob{\N_{t+\Delta}>n+1\;\vert\;\N_{t}=n}=o(\Delta),\\
    \Prob{\X_{t+\Delta}\in\mathcal{E}\;\vert\;\X_{t}=x, \N_{t+\Delta}-\N_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}+\frac{o(\Delta)}{\Delta}.
  \end{gathered}
\end{equation*}
\AAK{[Do we need the last term?]}
We will further assume that $\X_t$ is non-explosive, \ie, that $\Prob{\N_t<\infty}=1$ for all $t$.
\AAK{[Is this equivalent to non-explosivity? Or merely an implication?]}

@@

Without loss of generality, one can assume, as we do here, that the sample paths $t\mapsto\X_t(\omega)$ for $\omega\in\Omega$ are right-continuous with left limits (\cadlag).
In fact, all of the processes we will describe in this paper will be taken to be \cadlag, and we will frequently need to refer to the left-limit.
Accordingly, if $\rdm{Z}_t$ is any \cadlag\ random process, we define
\begin{equation*}
  \leftlim{\rdm{Z}}_t\colonequals
  \begin{cases}
    \lim_{t'\uparrow{t}}\rdm{Z}_t, & t>0,\\
    \rdm{Z}_0, & t=0.\\
  \end{cases}
\end{equation*}
Note that $\leftlim{\rdm{Z}}_t$ is thus left-continuous with right limits.

@@

The non-explosiveness assumption implies that a.s., for every $t$, there is a finite, increasing sequence of random event times, which we denote by $\event(\Hist_t)\colonequals\left(\rdm{T}_k\right)_{k=1}^{\rdm{K}_t}$;
the length, $\rdm{K}_t$, of this sequence is itself random.
However, conditional on $\Hist_t$, $\rdm{T}_k$ and $\rdm{K}_t$, together with the mark process $\U_t$, and the population process $\X_t$ are deterministic.
One can write down an explicit probability measure for $\Hist_t$ in terms of these:
\begin{equation}\label{eq:Hdens}
  \begin{aligned}
    \pi^{\Hist}(\dd{\Hist_t})
    =&p_{0}(X_{0})\,\dd{X_0}\,\prod_{k=1}^{K_{t}}{\left(\alpha_{U_{T_k}}\left(T_k,\leftlim{X}_{T_k},X_{T_k}\right)\,\dd{X_{T_k}}\,\dd{T_k}\right)}\\
    &\qquad\times\exp{\left(-\sum_{u}{\int_{0}^{t}{\int{\alpha_u(t',\leftlim{X}_{t'},x')\,\dd{x'}}\,\dd{t'}}}\right)}.
  \end{aligned}
\end{equation}

@@

Our goal in this paper is to probabilistically characterize how the genealogical relationships among lineages evolve through time.
Accordingly, we develop some notation for this purpose.
To begin with, we assign to each lineage a unique number $j\in\Zp$.
This can be done in any fashion, so long as no two lineages ever receive the same number.
For example, when a new lineage arises, we can assign it the smallest integer that has not yet been assigned.
We will define the \emph{inventory process}, $\Inv_t$, so that, for every lineage $j\in\Zp$, $\Inv_t(j)$ is the deme in which $j$ is found.
However, when $t$ is before the birth or after the death of lineage $j$, then clearly $\Inv_t(j)\notin\Demes$.
We say in this case that lineage $j$ is in the \emph{underdeme}, which we denote using the symbol $\eth$, so that we can write $\Inv_t(j)=\eth$.
We define $\Demesplus\colonequals\Demes\cup\{\eth\}$ so that $\Inv:\Rp\times\Zp\to\Demesplus$.

@@

Different kinds of events that occur for the population process can have different kinds of effects on the inventory process, and indeed not every jump affects $\Inv_t$ at all.
From the point of view of the inventory process, there are five distinct \emph{pure types} of jumps, which we enumerate here.
\begin{enumerate}[(a)]
\item Birth-type events result in the branching of one or more new lineages, each from some existing lineage.
  If $j$ is one of the new lineages, we use the expression $\Anc(j)$ to refer to its ancestor.
  Examples of birth-type events include transmission events, speciations, and actual births.
  It is not assumed that all new lineages arising from a birth event share the same ancestor.
\item Death-type events result in the extinction of one or more lineages.
  Examples include recovery, death of a host, and species extinctions.
\item Migration-type events result in the movement of a lineage from one deme to another.
  Spatial movements, changes in host age or behavior, and progression of an infection can all be represented as migration-type events.
\item Sample-type events result in the collection of a sample from a lineage but do not in themselves affect the inventory process.
\item Neutral-type events result in no change to any of the lineages.
\end{enumerate}
\Cref{fig:markov_state} depicts an example with all five of the pure types.
It is not necessary that a jump fall into just one of these types.
It is allowable, for instance to have compound jumps that fall into more than one category.
For example, sample/death-type events, in which a lineage is simultaneously sampled and removed, have been used, as have birth/death events in which one lineage reproduces at the same moment that another dies.
The theory presented here places no restrictions on the complexity of the events that can occur.

However, we do impose the restriction that the \emph{production}, \ie the deme-specific number of lineages emerging from the event, be constant for all jumps of a given mark.
To be precise, the \emph{production} is defined to be a function $r:\Jumps\times\Demes\to\Zp$, such that $r^u_i$ lineages of deme $i$ emerge from each event of mark $u$.
We write $r^u=(r^u_i)_{i\in\Demes}$.
Note that we lineages that die as a result of an event do not count in the production.
Also, it is important to note that the parent lineage or lineages, if they survive the event, are always counted in the production.

Because different kinds of events may differ not only in the number of offspring they engender, but also in the number of parent lineages, and the distribution of offspring among parents and demes, there is implicitly a deterministic indicator function $Q_u$, for $u\in\Jumps$, (described below) that captures these properties.

@@

The structure of the state space for the inventory process, $\Inv_t$, has already been described.
It remains to define its stochastic dynamics.
The $\Inv_t$ process is driven by the population process $\X_t$ in that jumps in $\Inv_t$ do not occur except when jumps in $\X_t$ occur:
$\leftlim{\Inv}_t\ne\Inv_t$ implies $\leftlim{\N}_t\ne\N_t$.

At jumps of birth type and mark $u$, the appropriate number of random parents are selected from the appropriate deme(s) and each one sires the appropriate number of offspring in each deme.
At jumps of death type, the appropriate number of lineages are selected from the appropriate demes and moved to the underdeme.
At jumps of migration type, randomly selected lineages are moved between demes as appropriate.
At sample-type and neutral-type jumps, no change occurs.
Jumps of compound type are handled in the obvious way.

Crucially, the assumption that the population process $\X_t$ is Markov empowers us to assume that the individual lineages within each deme are exchangeable with respect to the inventory process.
Since the law governing $\X_t$ is independent of the identities of the individual lineages, and since the individual lineages are exchangeable by assumption, it follows that, conditional on $\Hist_t$, the jumps of the inventory are independent of one another.

@@

\citet{King2022}, showed how an unstructured population process induces a process on the space of genealogies.
Although we now treat a more general case, the construction is much the same, so we abbreviate the presentation.
Readers wishing for more detail should consult the earlier paper.

For our purposes, a \emph{genealogy} is a labeled, time-calibrated tree.
Its edges represent relationships of ancestry and descent among its nodes.
There are three distinct kinds of nodes:
\begin{inparaenum}[(i)]
\item \emph{tip nodes}, which represent labeled extant lineages;
\item \emph{internal nodes}, which represent ancestral events;
\item \emph{sample nodes}, which represent labeled samples.
\end{inparaenum}
More formally, we can take a genealogy, $G$, to be a finite sequence of internal nodes, together with a time.
Given $G$, the time is denoted $T(G)$; this is the time corresponding to the extant lineages.
The number of nodes is $K(G)$.
We order the nodes temporally, and denote the $k$-th node $\Node{p}_k(G)$ and we write $\Node{p}\in{G}$ if $\Node{p}$ is one of the nodes of $G$.
Each $\Node{p}\in\Gen$ has a creation-time, $T(\Node{p})$, a parent, $\Anc(\Node{p})$, and a deme.
We let $\event(G)$ denote the sequence of node times:
$\event(G)=\left(\Node{p}_k(G)\right)_{k=1}^{K(G)}$, where $0=T(\Node{p}_1(G)){\le}T(\Node{p}_2(G)){\le}\dots{\le}T(\Node{p}_{K(G)}(G)){\le}T(G)$.
%% The nodes are ordered so that $k<k'$ implies that $T(\Node{p}_k(\Gen))\le{T(\Node{p}_{k'}(\Gen))}$.
Root nodes are distinguished by being their own parents:
$\Node{p}$ is a root if and only if $\Anc(\Node{p})=\Node{p}$.
Every node also has one or more descendant nodes called \emph{children}.

@@

The population and inventory processes together induce a stochastic process, $\Gen_t$, on the space of genealogies.
In particular, at each event in the population process, one or more of the following changes happen to the genealogy, according to the type of the event:
\begin{enumerate}[(a)]
\item A birth-type event at time $t$ results in the creation of one new internal node for each parent lineage.
  In particular, if $j$ is one of the parent lineages, and $\Node{p}$ is the new node, then $T(\Node{p})=t$, $\Anc(\Node{p})$ is the node that was parent to $j$ prior to the event.
  The children of $\Node{p}$ include one new tip node for each new lineage sired by $j$, as well as $j$ itself.
\item In a death-type event, all the lineages $j$ that die are removed.
  Internal nodes without children are then recursively removed.
  Sample nodes are never removed.
\item In a migration-type event, one node is added for each migrating lineage;
  each one takes the migrating lineage as child.
  The ancestor of the new node is that which was parent to the migrating lineage prior to the event.
  The deme of the lineage changes accordingly.
\item At a sample-type event, one new sample node is introduced for each sampled lineage.
  Each one takes the sampled lineage as child.
  The ancestor of the sample node is that which was parent to the sampled lineage before the vent.
\item At a neutral-type event, no change is made to the genealogy.
\item Finally, events of compound type are accommodated by combining the foregoing rules.
\end{enumerate}
When an event results in the addition of one or more new nodes to a genealogy, the lineages which are children of that node are said to \emph{emerge} from the event.
Thus, after a birth-type event, the emerging lineages include all the new offspring as well as their parents.
Likewise, at pure migration- or sample-type events, each migrating or sampled lineage emerges from the event.
At pure death-type events, no lineages emerge.
In general, at an event of mark $u$, there are $r^u_i$ emergent lineages in deme $i$.

@@

Although the process just described yields a genealogy that relates all extant members of the population, and all samples, the data we ultimately wish to analyze will contain only the samples.
We therefore describe how the genealogy process is \emph{pruned} to yield the sample-only genealogy.
Given a genealogy $\Gen_t$, one obtains the \emph{pruned genealogy}, $\Prune_t=\prune(\Gen_t)$ by first dropping every tip node and then recursively dropping every internal node without children.
In a pruned genealogy only internal and sample nodes remain, and sample nodes are found at all of the leaves and some of the interior nodes of the genealogy.
Observe that the pruned genealogy retains information not only about how much ancestry is shared by any pair of sample lineages, but also about where among the demes each lineage was through time.
Note also that $T(\Prune_t)=T(\Gen_t)=t$.

\paragraph{Alternative representation of a pruned genealogy}

Consider the finite set of all samples represented in pruned genealogy $P$.
Beginning with 1, assign natural numbers to these in such a way that if sample $j$ is ancestral to sample $j'$, then $j<j'$.
For example, we can order the samples by their times, resolving any ties arbitrarily.
Let $\Lin(P)$ denote the set of lineage numbers.
Using this ordering, we can uniquely associate each point on a genealogical tree with the least of those lineages that descend from that point.
In particular, any lineage $j\in\Lin(P)$, corresponding to a sample taken at time $t^s_j$, can be traced backward from node to node until either it coalesces with some lesser (\ie older) lineage at some time $t^o_j>0$ or a root is reached (in which case, we define $t^o_j=0$).
Each node encountered along the way represents a genealogical event from which $j$ emerges.
Moreover, at each time $t\in\halfopen{t^o_j,t^s_j}$, lineage $j$ is in precisely one of the demes $\Demes$.
However, for $t\notin\halfopen{t^o_j,t^s_j}$, lineage $j$ does not exist.
To express this, we again say that lineage $j$ is in the underdeme, $\eth$.

It will be useful to distill the elements that characterize a pruned genealogy.
Accordingly, given a pruned genealogy, $P$, we make the following definitions.
\begin{compactenum}[(a)]
\item Let $\ct^{P}:\Zp\times\Rp\to\Zp$ be such that, for $j\in\Zp$, $\ct^{P}_j(t)$ is the counting process which increases by 1 at each event along lineage $j$.
\item Let $\anc^{P}:\Zp\times\Rp\to\Zp$ be such that $\anc^{P}_j(t)$ indicates the unique lineage, ancestral of lineage $j$, and alive at time $t$.
  In particular, we posit that $\anc^{P}_j(t)=j$ for $t\in\halfopen{t^o_j,t^s_j}$ and $\anc^{P}_j(t)=0$ for $t>t^s_j$, so that the function is well defined for all $j$ and $t$.
  %%  Alternatively, one can define $\anc$ using the original construction of Kingman's coalescent \citep{Kingman1982a}.
\item Let $\deme^{P}:\Zp\times\Rp\to\Demesplus$ be such that, for $j\in\Zp$, $\deme^{P}_j(t)$ indicates in which deme lineage $j$ lies at time $t$.
  In particular $\deme^{P}_j(t)=\eth$ if $t\notin\halfopen{t^o_j,t^s_j}$.
\item Let $\Yspace=(\Zp\times\Zp\times\Demesplus)^{\Zp}$ and define $Y^{P}:\Rp\to\Yspace$ such that, for $j\in\Zp$ and $t\in\Rp$, $Y^{P}_j(t)=\left(\ct^{P}_j(t),\anc^{P}_j(t),\deme^{P}_j(t)\right)$.
\end{compactenum}
For $j>|\Lin(P)|$, we adopt the convention that $t^o_j=t^s_j=\infty$, so that $\ct^P_j(t)=0$, $\anc^{P}_j(t)=0$, and $\deme^P_j(t)=\eth$ for all $t\in\Rp$.
It is easy to see that $Y^{P}$ is well defined, piecewise constant, and \cadlag, and that the map $P\mapsto{Y^{P}}$ is one-to-one.
However, not every piecewise constant, \cadlag\ map $Y:\Rp\to\Yspace$ defines a pruned genealogy.

Given $\eta\in\Yspace$, we will use the notation $\ct(\eta)$, $\anc(\eta)$, and $\deme(\eta)$ to refer to the coordinates of $\eta$.

Since $\Gen_t$ is a stochastic process, both $\Prune_t\colonequals\prune(G_t)$ and $Y^{\Prune_t}$ are stochastic processes as well.
In fact, the latter two are Markovian, since each contains within itself all of its past history.
The process $\Gen_t$ is not Markovian, though $(\X_t,\Gen_t)$ is.

To visualize $\Prune_t$, one can make a correspondence between demes and colors.
Then a pruned genealogy is visualized as a tree with colored branches.
Knowing the function $\deme^{\Prune_t}$ is equivalent to knowing the coloring, while $\ct^{\Prune_t}$ determines the locations of events in the genealogy and $\anc^{\Prune_t}$ determines the topology.
Note in particular that, if $y=Y^{\Prune_t}$, then $y_j(t')\ne\yt_j(t')$ if and only if $t'$ is the time of an event from which lineage $j$ emerges.
Note also that, there can be events on a pruned genealogy where the color does not change.
That is, where $\ct_j$ increments, but neither $\anc_j$ nor $\deme_j$ change.

@@

In the following, we will find that we need to count the deme-specific numbers of lineages present at a given time.
Accordingly, for any $\eta\in\Yspace$, and $i\in\Demes$, let us define
\begin{equation*}
  \ell_i(\eta)\colonequals{\left\vert\left\{j\in\Zp\;\vert\;\deme(\eta_j)=i\right\}\right\vert}\in\Zp, \qquad \ell(\eta)\colonequals(\ell_i(\eta))_{i\in\Demes}\in{\Zp^\Demes}.
\end{equation*}
Note that lineages $j$ for which $\deme(\eta_j)=\eth$ are not counted.
With this definition, it follows that, for any pruned genealogy $P$, $\ell_i(Y^{P}(t))$ is the number of lineages in deme $i$ at time $t$ and
$\ell(Y^{P}(t))\in\Zp^\Demes$ is the non-negative integer vector telling how many lineages lie within each of the demes at time $t$.

We will also have occasion to refer to the deme-specific number of lineages emerging from a given event.
Therefore, for $\eta,\eta'\in\Yspace$, and $i\in\Demes$, let us define
\begin{equation*}
  s_i(\eta,\eta')\colonequals{\left\vert\left\{j\in{\Zp}\;\vert\;\deme(\eta_j)=i\ \ \&\ \ \ct(\eta'_j)\ne\ct(\eta_j)\right\}\right\vert}, \qquad s(\eta,\eta')\colonequals(s_i(\eta,\eta'))_{i\in\Demes}\in{\Zp^\Demes}.
\end{equation*}
With this definition, for any $P$, if $y=Y^P$, then $s_i(\yt(t),y(t))$ is the number of lineages in deme $i$ that emerge from an event at time $t$ and $s(\yt(t),y(t))$ is the non-negative integer vector telling how many lineages in each deme emerge at time $t$.
In particular, $s(\eta,\eta)=0$ so that if $t\notin\event(P)$, then $s(\yt(t),y(t))=0$.
We refer to $s\left(\yt(t),y(t)\right)$ as the \emph{saturation} at time $t$.

\subsection{Obscured genealogy}

As we have just seen, a pruned genealogy contains information about the full history of each sample lineage, including the times at which it entered or exited any deme, sired offspring, or was sampled.
The data we seek to analyze will typically lack much of this information.
Accordingly, we define the \emph{obscured genealogy} to be that obtained by discarding all information about demes and events not visible from the topology of the tree alone.
In particular, if $\Prune$ is a pruned genealogy and $\Vis=\obs(\Prune)$ is the corresponding obscured genealogy,
$\Vis$ is uniquely determined by the function $t\mapsto\anc(Y^{\Prune}(t))$.

\paragraph{Binomial ratio}

For $n,r,\ell,n\in{\Zp^\Demes}$, define the \emph{binomial ratio}
\begin{equation*}
  \BinRatio{n}{\ell}{r}{s}\colonequals
  \begin{cases}
    \frac{\prod_{i\in\Demes}{\binom{n_i-\ell_i}{r_i-s_i}}}{\prod_{i\in\Demes}{\binom{n_i}{r_i}}},
    & \forall i\ n_i\ge{\{\ell_i,r_i\}}\ge{s_i}\ge{0}\\
    0, & \text{otherwise}
  \end{cases}
\end{equation*}
Observe that $\BinRatio{n}{\ell}{r}{s}\in{[0,1]}$.
Moreover, in consequence of the Chu-Vandermonde identity, we have
\begin{equation*}
  \sum_{s\in\Zp^{\Demes}}\BinRatio{n}{\ell}{r}{s}\binom{\ell}{s}=1,
\end{equation*}
whenever $n_i\ge{\{\ell_i,r_i\}}\ge{0}$ for all $i$.

@@
